manuscript_v1

### 2022-01-31 manuscript.Rmd
manuscript v1.2 (intro complete)

---
title: "manuscript_v1"
author: "Karthik Thrikkadeeri"
date: "28/11/2021"
bibliography: msc-thesis.bib
output: 
  word_document:
    number_sections: true
---

@@

## Habitat selection in birds

The origins of the study of habitat selection in birds date back a considerable number of years (see Grinnell, 1917; Hildén, 1965; Svärdson, 1949; Block and Brennan, 1993; Cody, 1985). Its theoretical corpus has evolved from the early correlative models of habitat characteristics and species abundance (MacArthur and Pianka, 1966; Verner et al., 1986; Rosenzweig, 1991), to density dependence (Fretwell and Lucas, 1969; Fretwell, 1972) and more recently to broader multifaceted approaches. Despite the historical ambiguity (Hall et al., 1997) in the terms ‘habitat use’ and ‘habitat selection’, the current understanding is that habitat-use patterns are the end result of habitat-selection processes (Cody, 1985; Block and Brennan, 1993).

Animals select habitats by assessing the environment at multiple spatial scales and making a series of hierarchical choices (Johnson, 1980; Cody, 1985). Broad-scale selection is reflected in species’ geographic ranges and in the landscape features surrounding home ranges, while fine-scale selection is reflected in microhabitats used for foraging, reproduction, and shelter (Johnson, 1980; Saab, 1999). Species vary in their flexibility of habitat selection, from generalists to specialists, with specialists being less flexible and more susceptible to change (Devictor et al., 2008; Owens and Bennett, 2000). Intraspecific competition forces some individuals to occupy less suitable habitats, widening the ‘habitat breadth’ of that species, while interspecific competition has the opposite effect (Svärdson, 1949). Habitats chosen by animals can influence fitness outcomes via the costs (e.g., predation risk) and benefits (e.g., food availability) of habitat use.

Notably, the literature on habitat selection in birds, and more generally in free-ranging animals, has a number of biases, the most striking of which is that a majority targets migratory birds (e.g., Gailly et al., 2020; Jenkins et al., 2017; Szymkowiak and Thomson, 2019). This attention is not unjustified---migrants have complex habitat requirements that vary through the annual cycle, across their geographic range, along migratory routes, and at multiple spatial scales. Therefore, investigating their selection patterns may reveal more about the links between organisms and habitat and landscape features than those of resident species would. However, there are additional nested biases: most studies conducted in breeding grounds focus on the pre-breeding and breeding periods (e.g., Danchin et al., 1998; Kleist et al., 2017; Bosco et al., 2021), while some study wintering grounds (usually in the tropics; e.g., Strasser et al., 2019; Ruiz-Sanchez et al., 2017; Albert et al., 2020). To a lesser extent, there are studies concerning stopover sites which are equally important (e.g., Papageorgiou et al., 2017; Wolfe et al., 2014; Alonso et al., 2020; Cohen et al., 2021). There is also a significant body of literature concerning the dynamics of long-distance migrants like shorebirds and other large-bodied birds (e.g., Jourdan et al., 2021; Wang et al., 2019; Devries et al., 2018), which are relatively easier to track at the large scale.

This disproportionate attention given by researchers to breeding- and nest-site selection could be because the fitness costs and benefits resulting from selection have traditionally been linked with reproductive success and offspring survival. Nevertheless, habitat dynamics during transitional extra-breeding periods can be of great importance as well, given the associated high stress.

## Post-breeding season

In breeding grounds, habitat selection on the other end of the timeline, i.e., during late summer and early autumn prior to winter migration, remains understudied. Questions regarding community structure and niche breadth take on new dimensions in these periods. For one, territoriality is lower and mobility is higher, allowing birds greater freedom in choosing foraging areas (Uesugi and Murakami, 2007). Moreover, motivations for habitat selection are straightforward and unambiguous; the main challenges are food scarcity coupled with increased food requirements---both for migrants that need to build up enough nutritional reserves and for residents that need to prepare for the oncoming winter (via i.a., food-search and caching)---and the consequently higher predation risk.

The period of late summer and early autumn, prior to migration, also sees a transition from parent-dependent fledgelings to independent young birds that have to overcome challenges brought by the changing season and by their lack of experience in foraging, navigating and avoiding predators (Anders et al., 1997). Especially to them but also to adults, dense vegetation may provide cover and protection from adverse weather and predators. Furthermore, dispersing juveniles may also collect social information regarding future site quality during the post-breeding period (Betts et al., 2008a). Despite slowly growing evidence that habitat selection in birds may differ seasonally due to the varying ecological requirements during the different seasons (Rivera et al., 1998; King et al., 2006; Akresh et al., 2009; Vitz and Rodewald, 2011; Dittmar et al., 2014), this period remains understudied.

## The role of competition

In general, interspecific competition is likely to be higher among closely related species due to similar resource use patterns which would lead to smaller overall habitat breadths and overlaps, whereas without competitive effects related species can be expected to overlap more since they have similar basic feeding patterns. Alatalo (1981) found evidence for the prevalence of interspecific competition in the seasonal environment of Finland and noted that this ecological process might be working at two different time scales: either in current ecological time as exemplified by interspecific territoriality in some bird species (e.g., Orians and Willson, 1964) or during the evolution of feeding adaptations of the species---for example, tit species are better adapted to exploit resources in their preferred habitats than in other habitats (Partridge, 1976; Snow, 1954).

The role of competition as a driver of habitat use can be diverse and can vary with seasons. The competitive release hypothesis (Gauthreaux Jr, 1982; sensu Chapman et al., 2011), concerning specifically intraspecific competition, may just as well be applied to interspecific competition. It states that competition over limited resources in the breeding grounds during winter can predict patterns of partial migration; the decrease in overall resource availability is balanced by a reduction in the number of individuals that need resources. Therefore, species that are highly territorial and competitive in the breeding season might switch to cooperation due to competitive release, as suggested by the widespread occurrence of mixed-species foraging flocks (Greenberg, 2001). In closed habitats such as woodlands, predators attack at a close range and feeding competition is generally lower between species than within, thus favouring mixed associations of small numbers of several to many species (Terborgh, 1990). Risk of predation is thought to be a major factor driving the formation of mixed-species flocks, and species in mixed-species flocks show higher foraging and lower vigilance rates (Sridhar et al., 2009). In the absence of strong competition, birds can be expected to select for habitats that maximise foraging efficiency and, eventually, survival.

## Shifts in foraging strategy

Aside from competition, there are several other factors that may change with the seasons and that birds might select for. Migrating land birds exhibit a high level of foraging and dietary flexibility (Herrera, 1978; Gauthreaux Jr, 1982; Parrish, 1997, 2000; McWilliams and Karasov, 2001; Salewski and Jones, 2006). Species generally considered to be insectivorous, at least during the breeding period, may become facultatively frugivorous in autumn due to lower abundances of arthropods and greater availability and accessibility of fruits (Bairlein and Simons, 1995; Faaborg, 2002; Bairlein, 2003; Ottich and Dierschke, 2003; Newton, 2010; Hernández, 2009). Frugivory, due to the high concentrations of fatty acids in fruits, allows the birds to rapidly build up the fat reserves vital for migration. Within frugivory, birds may exhibit further selectivity by preferentially consuming fruits of native plants over those of invasive ones in late autumn (Smith et al., 2013), even if fruits of the latter are more abundant in this vital pre-migratory period (Gallinat et al., 2020). This is thought to be due to nutritional differences between the two (Ingold and Craycraft, 1983; Smith et al., 2013), although some studies have found no such differences (Greenberg and Walter, 2010) and this preference may also be reversed during winter (Greenberg and Walter, 2010).

It is worth noting an important distinction between temperate forests of the Old and New Worlds. Firstly, there are no specialised obligate frugivorous species in the Old World temperate avifauna. Moreover, European temperate forests have a much lower diversity of fruiting plants than American ones. Among the most widespread fruiting species are bramble *Rubus* spp., bilberry and blueberry *Vaccinium* spp., elderberry *Sambucus nigra*, rowan *Sorbus aucuparia* and alder buckthorn *Frangula alnus*. These species are typical of hedgerows, wood edges and disturbed patches, and old growth mixed forests, but are scarcely found in the large, compact and structurally poor stands of present-day Central European forests.

Contrarily to the migratory birds reliant on fruits, some wintering birds have been known to consume large numbers of terrestrial arthropods even in winter. Tits (Paridae) in spruce forests commonly consume Lepidoptera, Hemiptera, Coleoptera and spiders (Betts, 1955). For example, Gibb (1960) found that birds in winter consumed around 50% of the stock of the eucosmid Ernarmonia conicolana which is known to be an important food source of blue tits. Other observations from forests suggest that spiders and insects including eggs, larvae, pupae and adults in diapause are also important winter food sources for insectivorous songbirds (Williams and Batzli, 1979; Jansson and Brömssen, 1981; Laurent, 1986). There is also strong evidence that birds can indirectly impact terrestrial plant communities through arthropod predation (Van Bael et al., 2008; Mooney et al., 2010; Mäntylä et al., 2011), although few such studies in temperate regions have explored these effects in winter.

## Habitat features and quality

Canopy cover, tree density and richness, etc. are commonly used habitat characteristics that are compared with bird abundances or densities. Several studies have shown their links with bird preferences (Vitz and Rodewald, 2007). Modern forestry is a major factor that has been influencing landscape structure (Harris, 1984), and most modern-day forests in Central Europe have been restructured towards commercially exploited plantations dominated by spruce that form compact growths with a closed above-tree layer at maturity (Fanta, 2007). ***But many are highly heterogenous due to management.*** Spatial segregation of habitats at a relatively fine-grained scale is suggested to allow for the co-occurrence of more species. Berg (1997) found that bird abundances correlated positively with the presence of deciduous trees and with tree diameter, while among conifers the Norway spruce *Picea abies* was preferred over the pine *Pinus sylvestris*; total bird densities were also higher in fragments than in forest sites. Furthermore, birds might even show within-season movement up or down gradients of habitat quality. Such questions have scarcely been explored, but there have been affirmative findings at least in the breeding season (Betts et al., 2008b).

## Aims

In this study, I investigated patterns of habitat selection exhibited by post-breeding forest bird communities. Multiple smaller patches of varying vegetational characteristics were chosen within a single larger stretch of forest, and bird abundances were studied starting from late summer until early autumn, covering the post-breeding season at a fine-grained temporal resolution. The main objectives were to identify the habitat features that are tracked most prominently by forest birds prior to periods of high stress (i.e., autumn-winter), to assess the patterns and magnitude of habitat selection as well as possible temporal dynamism, and to explore the possibilities of ecological interactions and strategies.

@@

@@

@@

>>>>

---
title: 
author: "Karthik Thrikkadeeri"
date: "28/11/2021"
bibliography: msc-thesis.bib
output: 
  word_document:
    reference_docx: rmd_word_template.docx
    number_sections: true
editor_options: 
  chunk_output_type: console
---

@@

Much of our knowledge on habitat selection in free-ranging animals is restricted to the breeding season. At different scales, animals assess a wide variety of aspects and features of the environment and make a series of hierarchical choices, culminating in an affinity and preference for certain habitats over others, which is termed habitat selection *(Johnson, 1980; Cody, 1985)*. Studies on habitat selection are interested in the ways in which these preferences vary across species, space, time, etc. Broad-scale selection is reflected in species’ geographic ranges and in the landscape features surrounding home ranges, while fine-scale selection is reflected in microhabitats used for foraging, reproduction, and shelter (Johnson, 1980; Saab, 1999). However, the breeding season---encompassing both the actual reproductive period, as well as the period leading up to mating---dominates a large majority of these questions (e.g., Danchin et al., 1998; Kleist et al., 2017; Bosco et al., 2021). Moreover, most such studies target migratory species, whether it be in their breeding grounds (e.g., Danchin et al., 1998; Kleist et al., 2017; Bosco et al., 2021), wintering grounds (usually in the tropics; e.g., Strasser et al., 2019; Ruiz-Sanchez et al., 2017; Albert et al., 2020) or stopover sites (e.g., Papageorgiou et al., 2017; Wolfe et al., 2014; Alonso et al., 2020; Cohen et al., 2021).

This disproportionate attention is not unjustified: habitats chosen by animals can influence fitness outcomes via costs (e.g., predation risk) and benefits (e.g., food availability) of habitat use, and these have traditionally been strongly linked with reproductive success and offspring survival. In migrants, habitat requirements are complex and vary through the annual cycle, across their geographic range, along migratory routes, and at multiple spatial scales. Therefore, investigating their selection patterns may reveal a lot about the links between organisms and habitat and landscape features. Nevertheless, habitat dynamics during transitional extra-breeding periods can be of great importance as well, given their associated high stress and consequence as an important link in the chain. In the context of temperate birds, questions regarding habitat selection, niche breadth and community structure take on new dimensions in these periods. For one, the birds are less territorial and more mobile, allowing greater freedom in choosing foraging areas (Uesugi and Murakami, 2007). Additionally, motivations for habitat selection are straightforward and unambiguous; the main challenges are food scarcity coupled with increased food requirements, and the consequently higher predation risk---both for migrants that need to build up enough nutritional reserves and for residents that need to prepare for the oncoming winter. Parent-dependent fledgelings also transition to independent young birds that have to overcome challenges brought by the changing season and by their lack of experience in foraging, navigating and avoiding predators (Anders et al., 1997). Especially to them but also to adults, dense vegetation may provide cover and protection from adverse weather and predators. Dispersing juveniles may even collect social information regarding future site quality during the post-breeding period (Betts et al., 2008a). Despite slowly growing evidence that habitat selection in birds may differ seasonally due to the varying ecological requirements during the different seasons (Rivera et al., 1998; King et al., 2006; Akresh et al., 2009; Vitz and Rodewald, 2011; Dittmar et al., 2014), this period remains understudied.

Species vary in their degree of selectivity from generalists to specialists, with the latter being less flexible and more susceptible to change (Devictor et al., 2008; Owens and Bennett, 2000). Many migrating land birds generally considered to be insectivorous, at least during the breeding period, cope with the food pressure in the post-breeding period by exhibiting a high level of foraging and dietary flexibility (Herrera, 1978; Gauthreaux Jr, 1982; Parrish, 1997, 2000; McWilliams and Karasov, 2001; Salewski and Jones, 2006). These birds may become facultatively frugivorous in autumn, allowing them to rapidly build up the fat reserves vital for migration, as well as making them less reliant on the declining arthropods (Bairlein and Simons, 1995; Faaborg, 2002; Bairlein, 2003; Ottich and Dierschke, 2003; Newton, 2010; Hernández, 2009). However, an important distinction between temperate forests of the Old and New Worlds is that there are no specialised obligate frugivorous species in the Old World temperate avifauna, and more importantly, European temperate forests have a much lower diversity of fruiting plants than American ones. Among the most widespread fruiting species are bramble *Rubus* spp., bilberry and blueberry *Vaccinium* spp., elderberry *Sambucus nigra*, rowan *Sorbus aucuparia* and alder buckthorn *Frangula alnus*. These species are typical of hedgerows, wood edges and disturbed patches, and old growth mixed forests, but are scarcely found in the large, compact and structurally poor stands of present-day Central European forests. Therefore, the dietary options for Central European forest birds, particularly residents, are rather limited. Selection patterns are likely to be more dependent on structural characteristics of the habitat, such as heterogeneity and diversity, and whether the species chooses to adapt to varying features or to master one.

Oak forests are widely believed to be the most biodiversity-rich, supporting a highly diverse arthropod community which in turn supports a larger vertebrate community. These forests also have more heterogeneous canopies and understoreys than the structurally uniform coniferous forests. As a result, birds during the post-breeding period could show a strong foraging preference for oak forest patches. Terrestrial arthropods have been found to comprise a surprisingly large proportion of the diet of some wintering birds even in winter. For instance, Gibb (1960) found that birds in winter consumed around 50% of the stock of the eucosmid *Ernarmonia conicolana* which is known to be an important food source of blue tits. In spruce forests, tits (Paridae) commonly consume Lepidoptera, Hemiptera, Coleoptera and spiders (Betts, 1955). Other observations from forests suggest that spiders and insects including eggs, larvae, pupae and adults in diapause are also important winter food sources (Williams and Batzli, 1979; Jansson and Brömssen, 1981; Laurent, 1986). There is strong evidence that birds can indirectly impact terrestrial plant communities through arthropod predation (Van Bael et al., 2008; Mooney et al., 2010; Mäntylä et al., 2011), although few such studies in temperate regions have explored these effects in winter. Strictly insectivorous birds may be able to monopolise the arthropod food sources, while omnivores might enjoy the advantages that come with being generalists.

Further, under these various ecological pressures, competitive interactions may  arise. Intraspecific competition forces some individuals to occupy less suitable habitats, widening the habitat breadth of that species, while interspecific competition has the opposite effect (Svärdson, 1949). In the presence of interspecific competition, species with similar resource use patterns (such as those within the same guild) would have smaller overall habitat breadths and overlaps. On the other hand, the absence of competition would allow them to overlap more and take advantage of their shared feeding strategy. The competitive release hypothesis (Gauthreaux Jr, 1982; sensu Chapman et al., 2011) states that intraspecific competition over limited resources in the breeding grounds during winter can predict patterns of partial migration; the decrease in overall resource availability is balanced by a reduction in the number of individuals that need resources. When applied to interspecific competition, it suggests that species that are highly territorial and competitive in the breeding season might switch to cooperation due to competitive release, as evidenced by the widespread occurrence of mixed-species foraging flocks (Greenberg, 2001). In closed habitats such as woodlands, predators attack at a close range and feeding competition is generally lower between species than within, thus favouring mixed associations of small numbers of several to many species (Terborgh, 1990). Risk of predation is thought to be a major factor driving the formation of mixed-species flocks, and species in mixed-species flocks show higher foraging and lower vigilance rates (Sridhar et al., 2009). In the absence of strong competition, birds can be expected to select for habitats that maximise foraging efficiency and, eventually, survival. Thus, the role of competition as a driver of habitat selection can be diverse and highly variable.

These various strategies and interactions are usually linked with habitat characteristics such as canopy cover, heterogeneity, tree density and species richness (Vitz and Rodewald, 2007). Such characteristics are commonly compared with bird abundances or densities to analyse habitat selection. Berg (1997) found that bird abundances correlated positively with the presence of deciduous trees and with tree diameter, while among conifers the Norway spruce *Picea abies* was preferred over the pine *Pinus sylvestris*. Spatial segregation of habitats at a relatively fine-grained scale is suggested to allow for the co-occurrence of more species. However, in addition to hosting a low diversity of fruiting plant species, Central European forests are also less spatially heterogenous. Most modern-day forests in Central Europe have been restructured towards commercially exploited plantations dominated by spruce that form compact growths with a closed above-tree layer at maturity (Fanta, 2007). On the other hand, intense management and disturbance regimes bring in heterogeneity, but do not facilitate biodiversity for obvious reasons. In these circumstances, studies on habitat selection can provide crucial information that can better advise conservation action in the ever-globalising world.

In this study, we investigate patterns of habitat selection exhibited by temperate forest bird communities in the post-breeding period at a fine-grained temporal resolution. The main objectives are to identify the habitat features that are tracked most prominently in this period of high stress (i.e., the autumn-winter transition), to assess the patterns and magnitude of habitat selection as well as their possible temporal dynamism with the seasonal transition, and to explore the possibilities of ecological interactions and strategies.

@@

Multiple smaller patches of varying vegetational characteristics were chosen within a single larger stretch of forest, and bird abundances were studied starting from late summer until early autumn, covering the post-breeding season at a .

@@

 Alatalo (1981) found evidence for the prevalence of interspecific competition in the seasonal environment of Finland

@@

migratory birds joining resident ones could be responsible for some of the patterns. studies looking at proportion of winter community migrants is useful.
